6.829 Final Project: A comparison of bit rate selection algorithms

### 2013-05-17 paper/paper.tex
Added paragraph about what bit rates are.

\title{6.829 Final Project: A comparison of bit-rate selection algorithms}
\author{Colleen Josephson \\ \texttt{cjoseph@mit.edu}
  \and Pavel Panchekha \\ \texttt{pavpan@mit.edu}}
\date{May 2013}

@@

\begin{abstract}
We compare the performance SampleRate and Minstrel, two popular
bit-rate selection algorithms that have widespread real-world usage.
We use a trace-based approach to avoid kernel programming and consider
Python implementations of both SampleRate and Minstrel.  We test both
algorithms in multiple real-world scenarios, including scenarios with
mobile clients and noisy environments, to highlight differences
between the two algorithms.  We also introduce improvements to the
Minstrel algorithm that allow for significant gains in throughput.
\end{abstract}

@@

One of the key ways that wireless networks differ from wired links is
that wireless networks have varying link rates.  Link conditions vary
with time due to interference from other devices, changing network
geometry, and mobile clients.  An optimal rate at one time may be
different from the optimal rate just 30 seconds earlier.  A good
bit-rate selection algorithm has to detect and adapt to these
conditions.  If the chosen bit-rate is too slow, then the throughput
will be unnecessarily low; if the rate is too high, failures will be
very frequent and throughput will again suffer.

There are three main classes of bit-rate adaptation protocols:
frame-based, SNR-based, and cross-layer protocols.  Frame-based
protocols measure the fraction of successfully received packets.  SNR
protocols make decisions based on the estimated Signal to Noise ratio.
Cross-layer protocols use SoftPHY data from the physical layer.  The
most commonly implemented protocols on today's networks are frame
based, because SNR protocols perform poorly~\cite{samplerate}, and
cross-layer protocols cannot be deployed on current networks as they
violate the network layering abstraction.

The two most popular frame-based protocols are SampleRate and
Minstrel.  Both were implemented in the MadWifi drivers for Linux
wireless, and today Minstrel is the default bit-rate selection
algorithm for all wireless drivers on Linux.  Our research used a
trace-based approach to analyze the performance of these two
algorithms.  We also created a modified version of Minstrel that
provides significant throughput gains over the vanilla Minstrel
implementation.

This paper will provide a general overview of SampleRate and Minstrel.
We then discuss our testing methodology.  We compare the performance
of the two algorithms, and then discuss modifications we made to
Minstrel and compare it's performance to the current kernel
implementation.

\section{SampleRate}

SampleRate, first introduced in \cite{samplerate}, is a bit rate
selection algorithm which maintains estimates of average transmission
time for each potential bit rate.To keep these estimates up to date,
SampleRate periodically sends a packet at a randomly-selected bit
rate, and uses the success or failure of this packet to update the
average transmission time of that bit-rate.  SampleRate also abandons
a bit rate after four successive failures at that rate.  All
information about packets is maintained with a 10-second sliding window.

SampleRate has three main functions: \texttt{ApplyRate()}, which
returns a the bit-rate to send a packet at,
\texttt{ProcessFeedback()}, which updates the statistics for a
bit-rate after an attempted packet send, and
\texttt{RemoveStaleResults()}, which removes the effect of packets
older than 10 seconds.

\section{Minstrel}

@@

Minstrel implementation is based on SampleRate, so it also has
\texttt{ApplyRate()} and \texttt{ProcessFeedback()} methods, as well
as a \texttt{UpdateStats()} method that runs every 100ms.
\texttt{UpdateStats()} is home to some of the key differences between
Minstrel and SampleRate.  Instead of making decisions based on the
average transmission time, Minstrel uses throughput, computed with $$T
= \frac{p}{tx\_time}$$ where $T$ is the throughput of some rate $r$,
$p$ is the probability of successfully transmitting at $r$, and
$tx\_time$ is the computed lossless transmission time at rate $r$.

@@

\section{Methodology}

@@

Our initial plan was to use the algorithms as implemented in the
MadWifi drivers for the Atheros chip.  However, MadWifi is deprecated,
and only runs on now-ancient network cards.  In addition to the
unavailability of compatible hardware, MadWifi would not be an
accurate reflection of contemporary users.  Both ath5k and
\texttt{ath9k}, which were created to replace MadWifi, did not port
the old bit-rate selection system from MadWifi and instead opted to
use the Linux kernel's rate selection framework, which only implements
Minstrel.  Porting SampleRate over to the new drivers would not have
been trivial, as there are vast differences in the assumed interfaces.

@@

Traces were collected from an \texttt{ath9k} driver modified to sample
bit-rates uniformly (forgoing the usual bit rate selection algorithm)
and to report successes and failure from each packet.  Each bit-rate
was sampled uniformly in time -- thus, more packets were sent at
higher bit-rates, since those bit-rates otherwise deliver packets
faster.  For each packet sent, the time to send was recorded, and the
bit rate for the next packet was chosen to be the bit-rate that had
the smallest total transmissions time.  This guarantees approximately
uniform selection of bit-rates.  The wireless card used was an
Qualcomm Atheros \texttt{AR9285} Wireless Network Adapter.

@@

Eighteen total traces were collected, grouped into a number of
scenarios.  Each trace was 30 seconds to a minute long, and each
scenario was repeated twice.  One scenario placed the wireless card
with line of sight to the access point, two meters from it.  Another
created solid body interference by having one of the authors stand
between the wireless device and the access point.  One placed the
device around a corner from the access point, creating some potential
for multi-path interference and fading.  In one scenario, we moved the
wireless device throughout the trace, pacing back and forth in a 10
meter line in front of the access point.  Finally, one five-minute
trace was taken, which mixed the above scenarios, first moving around
a corner, then standing still, then moving to line-of-sight with the
access point.  A few shorter (3 to 15 second long) traces were
collected to investigate the startup behavior of bit-rate selection.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{bsmt13floorplan.png}
  \caption{We collected most traces at a remote access point in the basement of Building 13 at MIT.}
\label{figure:1}
\end{figure}

@@

Once collected, traces were fed to a bit rate selection simulator to
analyze the performance of Minstrel and SampleRate.  Both algorithms
were represented by two functions: \texttt{ApplyRate}, which returned
a multi-rate retry chain of rates to attempt transmission at, and
\texttt{ProcessFeedback}, which was passed the success or failure of
the transmission and how many attempts were made at each bit rate
(along with the current time and the total time to send the packet);
this parallels the interface used by the Linux kernel itself.  Helper
functions, like SampleRate's \texttt{RemoveStaleResults} function,
were instead called from one of \verb|apply_rate| or
\verb|process_feedback|.  For every packet transmission, the simulator
would call \texttt{ApplyRate}; compute the probability of success
based on packets sent in 10-millisecond window around the current
time; and from that determine the success or failure of the packet,
and the number of retries required.  The results of a test run are
thus non-deterministic; both algorithms tested are non-deterministic
anyway, so this is no handicap.  To improve re-implementation
fidelity, both implementations were meticulously checked by both
authors.

@@

\begin{figure*}[htbp]
  \centering
  \begin{subfigure}{.45\linewidth}
    \includegraphics[width=\linewidth]{constant.png}
    \caption{Throughput of using a constant 802.11b/g bit-rate in
      clear line-of-sight from the access point.}
    \label{figure:2}
  \end{subfigure}\hfill
  \begin{subfigure}{.45\linewidth}
    \includegraphics[width=\linewidth]{MinVSSam1.png}
    \caption{Throughput of Minstrel and SampleRate in various
      scenarios.}
    \label{figure:3}
  \end{subfigure}
  \caption{Comparing Minstrel and Samplerate}
\end{figure*}

In Figure~\ref{figure:2} we show the performance of constant bit-rates
on the \texttt{clear\_1.dat} trace, which is a 35 second trace in
clear view of the access point.  The optimal rate here was 18 Mbps,
which achieved an average throughput of about 15 Mbps.  The next
highest bit-rate, 24 Mbps, shows a large plummet in performance, down
to a throughput of 1.17 Mbps.  This steep drop is charateristic, and
is also talked about in the SampleRate thesis~\cite{samplerate}.  It
demonstrates just how important bit rate selection is---the optimal
bit rate has typically much higher throughput than the next-best rate,
but if the algorithm strays too high the results can be devastating.

@@

The histograms of which rates the algorithms decided to send at
provided some useful insight into their behavior.  Table~\ref{table:2}
is a histogram of a 35 second corner trace.  In this situation
Minstrel outperforms SampleRate.  The constant rate throughputs
demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at,
and that no packets were sucessful at any higher rates.  SampleRate
does not use these rates except for probe packets, which happens once
every ten seconds.  Minstrel, since it does not terminate transmission
after four successive failures, spends a lot of time sending probe
packets at these high rates, which are never successful.  The only
reason low rates have fewer probe packets is because Minstrel
explicitly places them second in the retry chain, and because the main
rate usually succeeds.  Despite the fact that SampleRate spends less
time probing high rates, it is less successful.  This is because
Minstrel send the most packets at the three most successful bit-rates.
SampleRate, however, spends too much time sampling low bit rates.
Additionally, it spends a lot of time sending at 5.5 Mbps.  We suspect
that this might be because the equation SampleRate uses to estimate
average transmission time needs to be tweaked.
%   TODO: why does SampleRate spend time at 5.5 Mbps???

@@

\section{Improvements to Minstrel}

@@

To achieve, this weighing, we first recast the usual EWMA algorithm.
The usual algorithm computes $$p \gets \alpha p + (1 - \alpha)
\frac{n_i}{d_i}),$$ where $p$ is the EWMA probability, $\alpha$ is a
weight factor equal to $0.75$ in Minstrel, and $n_i$ and $d_i$ are the
number of successful and total packets in the 100ms window.  We next
write this as $$p \gets \frac{\beta p + \frac{n_i}{d_i}}{\beta + 1},$$
where $\beta = \alpha / (1 - \alpha)$.  We can now change the factor
in front of $n_i / d_i$ to account for the total number of packets
sent.  In our case, we chose to compute $$p \gets \frac{\beta p + w_i
  \frac{n_i}{d_i}}{\beta + w_i},$$ where the weight $w_i$ is $d_i / (d
/ b)$, with $d$ the total number of packets ever send and $b$ the
total number of 100ms windows ever seen.  Then $d / b$ is the average
number of packets per window and $d_i / (d / b)$ is the ratio between
the current window and the average window.  Note that if the current
window is of average size, this is exactly equal to the usual EWMA
result.  Finally, we can simplify the above equation to $$S p \gets
\frac{\beta \frac{d}{b} p + S n_i}{\beta \frac{d}{b} + d_i},$$ where
$S$ is any constant.  This equation is usable in fixed-precision
arithmetic, making it simple to implement in the Linux kernel (which
prefers to avoid floating point computations for portability reasons).
Because packets are high rates rarely succeed, the average window size
is a few packets; thus, the main qualitative difference between our
balanced EWMA and the original Minstrel EWMA is that non-probe packets
at probe bit-rates are weighed extraordinarily heavily, and this fact
allows us to avoid multiple failing 100ms windows.

@@

We noted another curious fact about Minstrel: the kernel
implementation of Minstrel samples rates very frequently.  This is due
to a flag the kernel uses to track which probe bit-rates were actually
sampled.  In cases where the probe packets are at a lower bit-rate
than the current bit-rate, the probe packets might never be sent if
the current bit-rate succeeds.  However, the flagging seems to
misbehave in certain cases, causing the extremely aggressive probing
we see in some traces.  For example, in one of the traces where the
client was around a corner from the access point, Minstrel sent almost
half of its packets at probe frequencies.  We modified Minstrel to be
less aggressive with probe packets, decreasing the sampling frequency
to approximately 10\%.  This lead to more improvement, especially on
noisy traces such as the cases where the client was around a corner or
moving.

\begin{table*}[htb]
    \centering
    \begin{tabular}{l|l|l|l|l}

    \textbf{Rate in Mbps} & \textbf{SampleRate} &
    \textbf{Minstrel} & \textbf{Minproved} &
    \textbf{Constant Rate Throughput}\\ \hline

    1   & 631  & 39    & 29    & 0.639 Mbps \\
    2   & 863  & 82    & 20    & 1.425 Mbps \\
    5.5 & 4814 & 110   & 805   & 4.406 Mbps \\
    6   & 423  & 137   & 276   & 4.603 Mbps \\
    9   & 553  & 327   & 57    & 4.630 Mbps \\
    11  & 3326 & 4121  & 3356  & 9.627 Mbps \\
    12  & 4649 & 9515  & 14285 & 9.444 Mbps \\
    18  & 1154 & 13295 & 15795 & 8.458 Mbps \\
    24  & 16   & 6909  & 1918  & 0 Mbps \\
    36  & 16   & 6944  & 2030  & 0 Mbps \\
    48  & 16   & 6888  & 2009  & 0 Mbps \\
    54  & 16   & 7308  & 1953  & 0 Mbps \\ \hline

@@

Overall, our improved Minstrel, which we call Minproved, often
out-performs Minstrel by multiple Mbps and never doing any worse; and
usually surpassing the best constant bit rate.  In only one case did
SampleRate achieve higher throughput than Minproved, and in this case
both surpassed the best constant bit-rate.  In cases where Minstrel
out-performed SampleRate, the same behavior was true of Minproved.  In
most cases, Minproved was $30\%$ faster than Minstrel, and in some the
improvement as great as $60\%$.

However, Minproved still makes many poor choices.  While a single
successful probe packet does not cause many hundreds of milliseconds
of failed packets, as it does in Minstrel, the single successful
packet can still cause one window's worth of failed packets.  Instead,
it would be best if a successful probe packet at a rarely-successful
bit-rate was treated more carefully.  One can imagine an algorithm,
for example, which would treat a single successful probe packet as a
reason to send a dozen more probe packets at that bit-rate, but does
not yet commit to sending hundreds of packets at that bit-rate.

@@

\section{Availability}

@@

\section{Future Work}

Very little analysis has been done of the performance of bit-rate
selection algorithms on 802.11n networks.  802.11n introduces many new
rates, as well as multiple-input/multiple output (MIMO) capabilities.
There is an 802.11n version of Minstrel, but there is no similar
implementation of SampleRate.

@@

\section{Conclusion}

@@

\section{Acknowledgments}

>>>>

\title{6.829 Final Project: A comparison of bit rate selection algorithms}
\author{Colleen Josephson \\ \texttt{cjoseph@mit.edu}
  \and Pavel Panchekha \\ \texttt{pavpan@mit.edu}}
\date{May 2013}

@@

\begin{abstract}
We compare the performance SampleRate and Minstrel, two popular
bit rate selection algorithms that have widespread real-world usage.
We use a trace-based approach to avoid kernel programming and consider
Python implementations of both SampleRate and Minstrel.  We test both
algorithms in multiple real-world scenarios, including scenarios with
mobile clients and noisy environments, to highlight differences
between the two algorithms.  We also introduce improvements to the
Minstrel algorithm that allow for significant gains in throughput.
\end{abstract}

@@

One of the key ways that wireless networks differ from wired links is
that wireless networks have varying link rates.  Link conditions vary
with time due to interference from other devices, changing network
geometry, and mobile clients.  An optimal rate at one time may be
different from the optimal rate just 30 seconds earlier.  A good
bit rate selection algorithm has to detect and adapt to these
conditions.  If the chosen bit rate is too slow, then the throughput
will be unnecessarily low; if the rate is too high, failures will be
very frequent and throughput will again suffer.

The wireless 802.11 standard includes a variety of different bit rates
to account for a range of possible network speeds.  Later standard
versions introduce more bit rates; we focus on bit rates available in
802.11b and 802.11g, since 802.11n networks and hardware were not
available for study.  Table~\ref{tbl:bitrates} lists the bit rates
available in the wireless networks studied.  The job of a bit rate
selection protocol is to choose a bit rate, from among the available
rates, for each packet transmission.  Due to the variability and
non-determinism of packet transmission, this is a challenging task.

\begin{table}[bthp]
  \begin{tabular}{r|l|l}
    Throughput & Transmission & Modulation \\\hline

    1   Mbps & dsss & bpsk \\
    2   Mbps & dsss & qpsk \\
    5.5 Mbps & dsss & cck \\
    11  Mbps & dsss & cck \\\hline

    6   Mbps & ofdm & bpsk \\
    9   Mbps & ofdm & bpsk \\
    12  Mbps & ofdm & qpsk \\
    18  Mbps & ofdm & qpsk \\
    24  Mbps & ofdm & qam-16 \\
    36  Mbps & ofdm & qam-16 \\
    48  Mbps & ofdm & qam-64 \\
    54  Mbps & ofdm & qam-64 \\
  \end{tabular}
  \caption{Bit rates available in 802.11b and 802.11g.  The first
    block, of four bit rates, contains bit rates introduced in
    802.11b, while the last eight bit rates were introduced in
    802.11g.}
  \label{tbl:bitrates}
\end{table}

There are three main classes of bit rate selection protocols:
frame-based, SNR-based, and cross-layer protocols.  Frame-based
protocols measure the fraction of successfully received packets.  SNR
protocols make decisions based on the estimated Signal to Noise ratio.
Cross-layer protocols use SoftPHY data from the physical layer to make
better estimates of bit rate properties.  The most commonly
implemented protocols on today's networks are frame based; SNR
protocols perform poorly~\cite{samplerate} and cross-layer protocols
are hard to deploy due to their violation the standard network
layering abstraction.  Due to their ubiquity, we analyze frame-based
protocols.

The two most popular frame-based protocols are SampleRate and
Minstrel.  Both were implemented in the MadWifi drivers for Linux
wireless, and today Minstrel is the default bit rate selection
algorithm for all wireless drivers on Linux.  We use a trace-based
approach to analyze the performance of these two algorithms.  We also
created a modified version of Minstrel that provides significant
throughput gains over the vanilla Minstrel implementation.

We provide a high-level overview of SampleRate and Minstrel in
Sections~\ref{sec:samplerate} and~\ref{sec:minstrel}.  Then we discuss
our testing methodology in Section~\ref{sec:methodology} and analyze
the results in Section~\ref{sec:analysis}.  Finally we introduce our
modifications to the Minstrel algorithm in Section~\ref{sec:minproved}
and analyze the performance of our improvements.

This paper will provide a general overview of SampleRate and Minstrel.
We then discuss our testing methodology.  We compare the performance
of the two algorithms, and then discuss modifications we made to
Minstrel and compare its performance to the current kernel
implementation.

\section{SampleRate} \label{sec:samplerate}

SampleRate, first introduced in \cite{samplerate}, is a bit rate
selection algorithm which maintains estimates of average transmission
time for each potential bit rate.To keep these estimates up to date,
SampleRate periodically sends a packet at a randomly-selected bit
rate, and uses the success or failure of this packet to update the
average transmission time of that bit rate.  SampleRate also abandons
a bit rate after four successive failures at that rate.  All
information about packets is maintained with a 10-second sliding window.

SampleRate has three main functions: \texttt{ApplyRate()}, which
returns a the bit rate to send a packet at,
\texttt{ProcessFeedback()}, which updates the statistics for a
bit rate after an attempted packet send, and
\texttt{RemoveStaleResults()}, which removes the effect of packets
older than 10 seconds.

\section{Minstrel} \label{sec:minstrel}

@@

Minstrel implementation is based on SampleRate, so it also has
\texttt{ApplyRate()} and \texttt{ProcessFeedback()} methods, as well
as a \texttt{UpdateStats()} method that runs every 100ms.
\texttt{UpdateStats()} is home to some of the key differences between
Minstrel and SampleRate.  Instead of making decisions based on the
average transmission time, Minstrel uses throughput, computed with $$T
= \frac{p}{\mathtt{tx\_time}}$$ where $T$ is the throughput of some
rate $r$, $p$ is the probability of successfully transmitting at $r$,
and $tx\_time$ is the computed lossless transmission time at rate $r$.

@@

\section{Methodology} \label{sec:methodology}

@@

Our initial plan was to use the algorithms as implemented in the
MadWifi drivers for the Atheros chip.  However, MadWifi is deprecated,
and only runs on now-ancient network cards.  In addition to the
unavailability of compatible hardware, MadWifi would not be an
accurate reflection of contemporary users.  Both ath5k and
\texttt{ath9k}, which were created to replace MadWifi, did not port
the old bit rate selection system from MadWifi and instead opted to
use the Linux kernel's rate selection framework, which only implements
Minstrel.  Porting SampleRate over to the new drivers would not have
been trivial, as there are vast differences in the assumed interfaces.

@@

Traces were collected from an \texttt{ath9k} driver modified to sample
bit rates uniformly (forgoing the usual bit rate selection algorithm)
and to report successes and failure from each packet.  Each bit rate
was sampled uniformly in time -- thus, more packets were sent at
higher bit rates, since those bit rates otherwise deliver packets
faster.  For each packet sent, the time to send was recorded, and the
bit rate for the next packet was chosen to be the bit rate that had
the smallest total transmissions time.  This guarantees approximately
uniform selection of bit rates.  The wireless card used was an
Qualcomm Atheros \texttt{AR9285} Wireless Network Adapter.

@@

Eighteen total traces were collected, grouped into a number of
scenarios.  Each trace was 30 seconds to a minute long, and each
scenario was repeated twice.  One scenario placed the wireless card
with line of sight to the access point, two meters from it.  Another
created solid body interference by having one of the authors stand
between the wireless device and the access point.  One placed the
device around a corner from the access point, creating some potential
for multi-path interference and fading.  In one scenario, we moved the
wireless device throughout the trace, pacing back and forth in a 10
meter line in front of the access point.  Finally, one five-minute
trace was taken, which mixed the above scenarios, first moving around
a corner, then standing still, then moving to line-of-sight with the
access point.  A few shorter (3 to 15 second long) traces were
collected to investigate the startup behavior of bit rate selection.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{bsmt13floorplan.png}
  \caption{We collected most traces at a remote access point in the
    basement of MIT's Building 13.}
\label{figure:1}
\end{figure}

@@

Once collected, traces were fed to a bit rate selection simulator to
analyze the performance of Minstrel and SampleRate.  Both algorithms
were represented by two functions: \texttt{ApplyRate}, which returned
a multi-rate retry chain of rates to attempt transmission at, and
\texttt{ProcessFeedback}, which was passed the success or failure of
the transmission and how many attempts were made at each bit rate
(along with the current time and the total time to send the packet);
this parallels the interface used by the Linux kernel itself.  Helper
functions, like SampleRate's \texttt{RemoveStaleResults} function,
were instead called from one of \texttt{ApplyRate} or
\texttt{ProcessFeedback}.  For every packet transmission, the
simulator would call \texttt{ApplyRate}; compute the probability of
success based on packets sent in 10-millisecond window around the
current time; and from that determine the success or failure of the
packet, and the number of retries required.  The results of a test run
are thus non-deterministic; both algorithms tested are
non-deterministic anyway, so this is no handicap.  To improve
re-implementation fidelity, both implementations were meticulously
checked by both authors.

@@

\begin{figure*}[htbp]
  \centering
  \begin{subfigure}{.45\linewidth}
    \includegraphics[width=\linewidth]{constant.png}
    \caption{Throughput of using a constant 802.11b/g bit rate in
      clear line-of-sight from the access point.}
    \label{figure:2}
  \end{subfigure}\hfill
  \begin{subfigure}{.45\linewidth}
    \includegraphics[width=\linewidth]{MinVSSam1.png}
    \caption{Throughput of Minstrel and SampleRate in various
      scenarios.}
    \label{figure:3}
  \end{subfigure}
  \caption{Comparing Minstrel and Samplerate}
\end{figure*}

In Figure~\ref{figure:2} we show the performance of constant bit rates
on the \texttt{clear\_1.dat} trace, which is a 35 second trace in
clear view of the access point.  The optimal rate here was 18 Mbps,
which achieved an average throughput of about 15 Mbps.  The next
highest bit rate, 24 Mbps, shows a large plummet in performance, down
to a throughput of 1.17 Mbps.  This steep drop is charateristic, and
is also talked about in the SampleRate thesis~\cite{samplerate}.  It
demonstrates just how important bit rate selection is---the optimal
bit rate has typically much higher throughput than the next-best rate,
but if the algorithm strays too high the results can be devastating.

@@

The histograms of which rates the algorithms decided to send at
provided some useful insight into their behavior.  Table~\ref{table:2}
is a histogram of a 35 second corner trace.  In this situation
Minstrel outperforms SampleRate.  The constant rate throughputs
demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at,
and that no packets were sucessful at any higher rates.  SampleRate
does not use these rates except for probe packets, which happens once
every ten seconds.  Minstrel, since it does not terminate transmission
after four successive failures, spends a lot of time sending probe
packets at these high rates, which are never successful.  The only
reason low rates have fewer probe packets is because Minstrel
explicitly places them second in the retry chain, and because the main
rate usually succeeds.  Despite the fact that SampleRate spends less
time probing high rates, it is less successful.  This is because
Minstrel send the most packets at the three most successful bit rates.
SampleRate, however, spends too much time sampling low bit rates.
Additionally, it spends a lot of time sending at 5.5 Mbps.  We suspect
that this might be because the equation SampleRate uses to estimate
average transmission time needs to be tweaked.
%   TODO: why does SampleRate spend time at 5.5 Mbps???

@@

\section{Improvements to Minstrel} \label{sec:minproved}

@@

To achieve, this weighing, we first recast the usual EWMA algorithm.
The usual algorithm computes $$p \gets \alpha p + (1 - \alpha)
\frac{n_i}{d_i}),$$ where $p$ is the EWMA probability, $\alpha$ is a
weight factor equal to $0.75$ in Minstrel, and $n_i$ and $d_i$ are the
number of successful and total packets in the 100ms window.  We next
write this as $$p \gets \frac{\beta p + \frac{n_i}{d_i}}{\beta + 1},$$
where $\beta = \alpha / (1 - \alpha)$.  We can now change the factor
in front of $n_i / d_i$ to account for the total number of packets
sent.  In our case, we chose to compute $$p \gets \frac{\beta p + w_i
  \frac{n_i}{d_i}}{\beta + w_i},$$ where the weight $w_i$ is $d_i / (d
/ b)$, with $d$ the total number of packets ever send and $b$ the
total number of 100ms windows ever seen.  Then $d / b$ is the average
number of packets per window and $d_i / (d / b)$ is the ratio between
the current window and the average window.  Note that if the current
window is of average size, this is exactly equal to the usual EWMA
result.  Finally, we can simplify the above equation to $$S p \gets
\frac{\beta \frac{d}{b} S p + S n_i}{\beta \frac{d}{b} + d_i},$$ where
$S$ is any constant.  This equation is usable in fixed-point
arithmetic (where $S$ is the scaling factor for the fixed-point
representation), making it suitable for implementation in the Linux
kernel (which avoids floating point computations for
portability reasons).  Because packets are high rates rarely succeed,
the average window size is a few packets; thus, the main qualitative
difference between our balanced EWMA and the original Minstrel EWMA is
that non-probe packets at probe bit rates are weighed extraordinarily
heavily, and this fact allows us to avoid multiple failing 100ms
windows.

@@

We noted another curious fact about Minstrel: the kernel
implementation of Minstrel samples rates very frequently.  This is due
to a flag the kernel uses to track which probe bit rates were actually
sampled.  In cases where the probe packets are at a lower bit rate
than the current bit rate, the probe packets might never be sent if
the current bit rate succeeds.  However, the flagging seems to
misbehave in certain cases, causing the extremely aggressive probing
we see in some traces.  For example, in one of the traces where the
client was around a corner from the access point, Minstrel sent almost
half of its packets at probe frequencies.  We modified Minstrel to be
less aggressive with probe packets, decreasing the sampling frequency
to approximately 10\%.  This lead to more improvement, especially on
noisy traces such as the cases where the client was around a corner or
moving.

\begin{table*}[htb]
    \centering
    \begin{tabular}{r|rrr|r}

    \textbf{Maximum Throughput} & \textbf{SampleRate} &
    \textbf{Minstrel} & \textbf{Minproved} &
    \textbf{Static Throughput}\\ \hline

    1   Mbps & 631  pkts & 39    pkts & 29    pkts & 0.639 Mbps \\
    2   Mbps & 863  pkts & 82    pkts & 20    pkts & 1.425 Mbps \\
    5.5 Mbps & 4814 pkts & 110   pkts & 805   pkts & 4.406 Mbps \\
    6   Mbps & 423  pkts & 137   pkts & 276   pkts & 4.603 Mbps \\
    9   Mbps & 553  pkts & 327   pkts & 57    pkts & 4.630 Mbps \\
    11  Mbps & 3326 pkts & 4121  pkts & 3356  pkts & 9.627 Mbps \\
    12  Mbps & 4649 pkts & 9515  pkts & 14285 pkts & 9.444 Mbps \\
    18  Mbps & 1154 pkts & 13295 pkts & 15795 pkts & 8.458 Mbps \\
    24  Mbps & 16   pkts & 6909  pkts & 1918  pkts & 0 Mbps \\
    36  Mbps & 16   pkts & 6944  pkts & 2030  pkts & 0 Mbps \\
    48  Mbps & 16   pkts & 6888  pkts & 2009  pkts & 0 Mbps \\
    54  Mbps & 16   pkts & 7308  pkts & 1953  pkts & 0 Mbps \\ \hline

@@

Overall, our improved Minstrel, which we call Minproved, often
out-performs Minstrel by multiple Mbps and never doing any worse; and
usually surpassing the best constant bit rate.  In only one case did
SampleRate achieve higher throughput than Minproved, and in this case
both surpassed the best constant bit rate.  In cases where Minstrel
out-performed SampleRate, the same behavior was true of Minproved.  In
most cases, Minproved was $30\%$ faster than Minstrel, and in some the
improvement as great as $60\%$.

However, Minproved still makes many poor choices.  While a single
successful probe packet does not cause many hundreds of milliseconds
of failed packets, as it does in Minstrel, the single successful
packet can still cause one window's worth of failed packets.  Instead,
it would be best if a successful probe packet at a rarely-successful
bit rate was treated more carefully.  One can imagine an algorithm,
for example, which would treat a single successful probe packet as a
reason to send a dozen more probe packets at that bit rate, but does
not yet commit to sending hundreds of packets at that bit rate.

@@

\section{Availability} \label{sec:availability}

@@

\section{Future Work} \label{sec:future-work}

Very little analysis has been done of the performance of bit rate
selection algorithms on 802.11n networks.  802.11n introduces many new
rates, as well as multiple-input/multiple output (MIMO) capabilities.
There is an 802.11n version of Minstrel, but there is no similar
implementation of SampleRate.

@@

\section{Conclusion} \label{sec:conclusion}

@@

\section*{Acknowledgments} \label{sec:acknowledgments}
